\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2018-11-05 fully-abstract-comp/popl2019/paper.tex
Capitalize section headers.

\subsection{Overlay semantics}
% Describe the things added to the syntax
The overlay semantics \srccm{} for \trgcm{} views part of the memory as a built-in stack (Figure~\ref{fig:source-syntax}).
% Built-in stack
To this end, it adds a call stack and a free stack memory to the executable configurations of \trgcm{}.
The call stack is a list with all the stack frames that are currently inaccessible because they belong to previous calls.
Every stack frame contains encapsulated stack memory as well as the program point that execution is supposed to return to.
The free stack memory is the active part of the stack that has not been claimed by a call and thus can be used at this point of time.
% Stack pointer
In order to distinguish capabilities for the stack from the capabilities for the rest of the memory, \srccm{} adds stack pointers.
A stack pointer has a permission, range of authority, and current address, just like capabilities on \trgcm{}, but they are always linear.
% Return pointers
The final syntactic constructs added by \srccm{} are the code and data return pointers.
The data return pointer corresponds to some stack pointer (which in turn corresponds to a linear capability), and the code return pointer corresponds to some capability with read-execute permission.
Syntactically, the return capabilities contain just enough information to reconstruct what they correspond to on the underlying machine.
On \srccm{}, return pointers are generated by calls from the capabilities they correspond to on \trgcm{}, and they are turned back to the capabilities they correspond to upon return.
% Even though the return pointers both correspond to capabilities with read permission, neither can be used for reading. 
% Say, we allowed reads through a data return pointer, then that would correspond to reading from some stack memory that is encapsulated in a stack frame which would break LSE.

@@

\subsection{Well-formedness and reasonability}

@@

\subsection{Full abstraction}
All that is left before we state the full-abstraction theorem is to define how components are combined with contexts and executed, so that we can define contextual equivalence.

@@

\subsection{The logical relation}
Using these Kripke worlds as assumptions, we can then define when different \srccm{} and \trgcm{} entities are related: values, jump targets, memories, execution configurations, components etc.
The most important relations are summarised in the following table, where we mention the general form of the relations, what type of things they relate and extra conditions that some of them imply:\\
% latex hack stolen from https://tex.stackexchange.com/questions/78788/align-equations-over-multiple-tabular-rows
% Lau: I find it difficult to easily see what constitutes a row in this table.
% I have updated it to a table I find easier to read.
\newcolumntype{R}{>{$}r<{$}}
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{M}{R@{${}\in{}$}L}
\begin{tabular}{|M|c|p{4.8cm}|}
  \hline
  \multicolumn{2}{|c|}{General form} & Relates ... & and ...\\
  \hline
  \npair{(w_S,w_T)} & \lrv(W) & values (machine words) & safe to pass to adversarial code\\
  \npair{(w_S,w_T)} & \lrvtrusted(W) & values (machine words) & \\
  \npair{(\reg_S,\reg_T)}  &  \lrr(W) & register files & safe to pass to adversarial code\\
  \npair{(\reg_S,\reg_T)}  &  \lrrtrusted(W) & register files & \\
  \npair{\Phi_S,\Phi_T}  &  \lro & execution configurations & \\
  \npair{(w_S,w_T)}  &  \lre(W) & $\tjmp{}$ targets &\\
  \left(\arraycolsep=1pt\array{l}(w_{S,1},w_{S,2}),\\(w_{T,1},w_{T,2})\endarray\right)  &  \lrexj(W) & $\txjmp{}{}$ targets &\\
  \multicolumn{2}{|c|}{$\memSat{\ms_S,\stk,\ms_\stk,\ms_T}{W}$} & memory & satisfy the assumptions in $W$\\
  \hline
\end{tabular}\\
These relations are defined using a set of mutually recursive equations, with cyclicity resolved through another use of step-indexing.
For space reasons, we cannot show all of these definitions, but we will try to give an overview.

@@

\subsection{Fundamental theorem}
An important lemma in our proof of full abstraction of the embedding of \srccm{} into \trgcm{}, is the fundamental theorem of logical relations (FTLR).
The name indicates that it is an instance of a general pattern in logical relations proofs, but is otherwise unimportant.
\begin{lemma}[FTLR (roughly)]
  \label{thm:ftlr}
  If $[\baddr,\eaddr] \subseteq \dom(\mscode)$ and $\pwheap(r) = \codereg{\sigrets,\sigcloss,\mscode}$, and either $[\baddr,\eaddr] \subseteq \ta$ and $\mscode$ behaves reasonably (see Section~\ref{sec:well-form-reas}) or
$[\baddr,\eaddr] \mathrel{\#} \ta$,
then 
  \[
    \npair{\left(\src{((\rx,\normal),\baddr,\eaddr,\aaddr)}, ((\rx,\normal),\baddr,\eaddr,\aaddr)\right)} \in \lre(W) \qedhere
  \]
\end{lemma}

@@

\subsection{Full abstraction proof sketch}
\label{subsec:proof-sketch}
Using Lemma~\ref{thm:ftlr}, we can now proceed to proving Theorem~\ref{thm:full-abstraction} (full abstraction).
First, we extend the logical relation into an omitted relation on components $(\var{comp}_S,\var{comp}_T) \in \lrcomp(W)$.
% Essentially, it relates a component to itself if instantiating their imports with related values produces related exports and code memory satisfying the appropriate code region in the world.
% The component relation $\mathcal{C}(W)$ basically lifts the logical relation we have presented above to components.
% The component relation relates a component $(\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss)$ to itself when two conditions are satisfied.
% First, when words that relate to them selves in the untrusted value relation are used to satisfy the imports, i.e.\ the words are placed on the import addresses in $\msdata$, and this data memory is combined with the code memory $\mscode$, then it forms a safe heap, i.e.\ it is in the $\mathcal{H}$ relation.
% Second, the exports should always be safe to use which means that they must be in the untrusted value relation in any future world.
% \begin{multline*}
%   \lrcomp(W) =\\
%   \left\{\begin{aligned}
%       &\npair{\var{comp},\var{comp}} \;\mid \;\\
%       &\qquad\var{comp} = (\mscode,\msdata,\overline{a_{\mathrm{import}} \mapsfrom s_{\mathrm{import}}},\overline{s_{\mathrm{export}} \mapsto w_{\mathrm{export}}},\sigrets,\sigcloss) \tand \\
%       &\qquad\text{For all } W' \future W \ldotp \\
%       &\qquad\quad\text{If } \overline{\npair[n']{(w_{\mathrm{import}},w_{\mathrm{import}})}} \in \lrv(\purePart{W'}) \text{ for all $n' < n$}\\
%       &\qquad\quad\text{and } \msdata' = \msdata{}[\overline{a_{\mathrm{import}} \mapsto w_{\mathrm{import}}}] \\
%       &\qquad\quad\text{then } \npair{(\sigrets\uplus\sigcloss,\mscode\uplus \msdata', \mscode\uplus\msdata')} \in \lrheap(\pwheap)(W') \tand\\
%       &\qquad\quad\overline{\npair{(w_{\mathrm{export}},w_{\mathrm{export}})}} \in \lrv(\purePart{W'})
%     \end{aligned}
%   \right\}\\
% \cup \left\{
%     \begin{multlined}
%       \npair{(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}),(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})} \;\mid \;\\
%       \npair{(\var{comp}_0,\var{comp}_0)} \in \lrcomp(W) \tand
%       \{(\_ \mapsto c_{\mathrm{main},c}),(\_ \mapsto c_{\mathrm{main},d})\} \subseteq \overline{w_{\mathrm{export}}}
%     \end{multlined}
%   \right\} 
% \end{multline*}
%
Using Lemma~\ref{thm:ftlr} and the definitions of the logical relations, we can then prove the following two lemmas.
The first is a version of the FTLR for components, stating that all components are related to themselves if they are either (1) well-formed and untrusted or (2) well-formed, reasonable and trusted.
\begin{lemma}[FTLR for components]
  \label{lem:ftlr-comps}
  If $\comp$ is a well-formed component, i.e. $\wdjud{\comp}$ and either
    $\dom(\comp.\mscode) \subseteq \ta$ and $\src{\comp}$ is a reasonable component; or
    $\dom(\comp.\mscode) \mathrel{\#} \ta$,
  then there exists a $W$ such that
  $\npair{(\src{\comp},\comp)} \in \lrcomp(W)$.
\end{lemma}

@@

\section{Discussion}
\label{sec:discussion}
% \begin{itemize}
% \item explain how fully abstract overlay semantics could form one pass of a verified secure compiler.
% \item Sharing stack references accross component boundaries is supported
% \item Other notions of well-bracketedness (specifically one would be to allow different stacks)
% \end{itemize}
\subsection{Full abstraction}
% - Full abstraction proofs difficult
Our formulation of WBCF and LSE using a fully abstract overlay semantics has an important advantage with respect to others.
Imagine that you are implementing a fully abstract compiler for a high-level language, i.e.\ a secure compiler that enforces high-level abstractions when interacting with untrusted target-language components.
Such a compiler would need to perform many things and enforce other high-level properties than just WBCF and LSE.

@@

%\subsection{Sharing the stack}
% ?
\subsection{Practical applicability}
We believe there are good arguments for practical applicability of \stktokens{}.
The strong security guarantees are proven in a way that is reusable as part of a bigger proof of compiler security.
Its costs are
\begin{itemize}
\item a constant and limited amount of checks on every boundary crossing.
\item possibly a small memory overhead because stack frames must be of non-zero length
\end{itemize}
The main caveat is that we rely on the assumption that capability machines like CHERI can be extended with linear capabilities in an efficient way.

>>>>

\subsection{Overlay Semantics}
% Describe the things added to the syntax
The overlay semantics \srccm{} for \trgcm{} views part of the memory as a built-in stack (Figure~\ref{fig:source-syntax}).
% Built-in stack
To this end, it adds a call stack and a free stack memory to the executable configurations of \trgcm{}.
The call stack is a list with all the stack frames that are currently inaccessible because they belong to previous calls.
Every stack frame contains encapsulated stack memory as well as the program point that execution is supposed to return to.
The free stack memory is the active part of the stack that has not been claimed by a call and thus can be used at this point of time.
% Stack pointer
In order to distinguish capabilities for the stack from the capabilities for the rest of the memory, \srccm{} adds stack pointers.
A stack pointer has a permission, range of authority, and current address, just like capabilities on \trgcm{}, but they are always linear.
% Return pointers
The final syntactic constructs added by \srccm{} are the code and data return pointers.
The data return pointer corresponds to some stack pointer (which in turn corresponds to a linear capability), and the code return pointer corresponds to some capability with read-execute permission.
Syntactically, the return capabilities contain just enough information to reconstruct what they correspond to on the underlying machine.
On \srccm{}, return pointers are generated by calls from the capabilities they correspond to on \trgcm{}, and they are turned back to the capabilities they correspond to upon return.
% Even though the return pointers both correspond to capabilities with read permission, neither can be used for reading. 
% Say, we allowed reads through a data return pointer, then that would correspond to reading from some stack memory that is encapsulated in a stack frame which would break LSE.

@@

\subsection{Well-Formedness and Reasonability}

@@

\subsection{Full Abstraction}
All that is left before we state the full-abstraction theorem is to define how components are combined with contexts and executed, so that we can define contextual equivalence.

@@

\subsection{The Logical Relation}
Using these Kripke worlds as assumptions, we can then define when different \srccm{} and \trgcm{} entities are related: values, jump targets, memories, execution configurations, components etc.
The most important relations are summarised in the following table, where we mention the general form of the relations, what type of things they relate and extra conditions that some of them imply:\\
% latex hack stolen from https://tex.stackexchange.com/questions/78788/align-equations-over-multiple-tabular-rows
% Lau: I find it difficult to easily see what constitutes a row in this table.
% I have updated it to a table I find easier to read.
\newcolumntype{R}{>{$}r<{$}}
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{M}{R@{${}\in{}$}L}
\begin{tabular}{|M|c|p{4.8cm}|}
  \hline
  \multicolumn{2}{|c|}{General form} & Relates ... & and ...\\
  \hline
  \npair{(w_S,w_T)} & \lrv(W) & values (machine words) & safe to pass to adversarial code\\
  \npair{(w_S,w_T)} & \lrvtrusted(W) & values (machine words) & \\
  \npair{(\reg_S,\reg_T)}  &  \lrr(W) & register files & safe to pass to adversarial code\\
  \npair{(\reg_S,\reg_T)}  &  \lrrtrusted(W) & register files & \\
  \npair{\Phi_S,\Phi_T}  &  \lro & execution configurations & \\
  \npair{(w_S,w_T)}  &  \lre(W) & $\tjmp{}$ targets &\\
  \left(\arraycolsep=1pt\array{l}(w_{S,1},w_{S,2}),\\(w_{T,1},w_{T,2})\endarray\right)  &  \lrexj(W) & $\txjmp{}{}$ targets &\\
  \multicolumn{2}{|c|}{$\memSat{\ms_S,\stk,\ms_\stk,\ms_T}{W}$} & memory & satisfy the assumptions in $W$\\
  \hline
\end{tabular}\\
These relations are defined using a set of mutually recursive equations, with cyclicity resolved through another use of step-indexing.
For space reasons, we cannot show all of these definitions, but we will try to give an overview.

@@

\subsection{Fundamental Theorem}
An important lemma in our proof of full abstraction of the embedding of \srccm{} into \trgcm{}, is the fundamental theorem of logical relations (FTLR).
The name indicates that it is an instance of a general pattern in logical relations proofs, but is otherwise unimportant.
\begin{lemma}[FTLR (roughly)]
  \label{thm:ftlr}
  If $[\baddr,\eaddr] \subseteq \dom(\mscode)$ and $\pwheap(r) = \codereg{\sigrets,\sigcloss,\mscode}$, and either $[\baddr,\eaddr] \subseteq \ta$ and $\mscode$ behaves reasonably (see Section~\ref{sec:well-form-reas}) or
$[\baddr,\eaddr] \mathrel{\#} \ta$,
then 
  \[
    \npair{\left(\src{((\rx,\normal),\baddr,\eaddr,\aaddr)}, ((\rx,\normal),\baddr,\eaddr,\aaddr)\right)} \in \lre(W) \qedhere
  \]
\end{lemma}

@@

\subsection{Full Abstraction Proof Sketch}
\label{subsec:proof-sketch}
Using Lemma~\ref{thm:ftlr}, we can now proceed to proving Theorem~\ref{thm:full-abstraction} (full abstraction).
First, we extend the logical relation into an omitted relation on components $(\var{comp}_S,\var{comp}_T) \in \lrcomp(W)$.
% Essentially, it relates a component to itself if instantiating their imports with related values produces related exports and code memory satisfying the appropriate code region in the world.
% The component relation $\mathcal{C}(W)$ basically lifts the logical relation we have presented above to components.
% The component relation relates a component $(\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss)$ to itself when two conditions are satisfied.
% First, when words that relate to them selves in the untrusted value relation are used to satisfy the imports, i.e.\ the words are placed on the import addresses in $\msdata$, and this data memory is combined with the code memory $\mscode$, then it forms a safe heap, i.e.\ it is in the $\mathcal{H}$ relation.
% Second, the exports should always be safe to use which means that they must be in the untrusted value relation in any future world.
% \begin{multline*}
%   \lrcomp(W) =\\
%   \left\{\begin{aligned}
%       &\npair{\var{comp},\var{comp}} \;\mid \;\\
%       &\qquad\var{comp} = (\mscode,\msdata,\overline{a_{\mathrm{import}} \mapsfrom s_{\mathrm{import}}},\overline{s_{\mathrm{export}} \mapsto w_{\mathrm{export}}},\sigrets,\sigcloss) \tand \\
%       &\qquad\text{For all } W' \future W \ldotp \\
%       &\qquad\quad\text{If } \overline{\npair[n']{(w_{\mathrm{import}},w_{\mathrm{import}})}} \in \lrv(\purePart{W'}) \text{ for all $n' < n$}\\
%       &\qquad\quad\text{and } \msdata' = \msdata{}[\overline{a_{\mathrm{import}} \mapsto w_{\mathrm{import}}}] \\
%       &\qquad\quad\text{then } \npair{(\sigrets\uplus\sigcloss,\mscode\uplus \msdata', \mscode\uplus\msdata')} \in \lrheap(\pwheap)(W') \tand\\
%       &\qquad\quad\overline{\npair{(w_{\mathrm{export}},w_{\mathrm{export}})}} \in \lrv(\purePart{W'})
%     \end{aligned}
%   \right\}\\
% \cup \left\{
%     \begin{multlined}
%       \npair{(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}),(\var{comp}_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})} \;\mid \;\\
%       \npair{(\var{comp}_0,\var{comp}_0)} \in \lrcomp(W) \tand
%       \{(\_ \mapsto c_{\mathrm{main},c}),(\_ \mapsto c_{\mathrm{main},d})\} \subseteq \overline{w_{\mathrm{export}}}
%     \end{multlined}
%   \right\} 
% \end{multline*}
%
Using Lemma~\ref{thm:ftlr} and the definitions of the logical relations, we can then prove the following two lemmas.
The first is a version of the FTLR for components, stating that all components are related to themselves if they are either (1) well-formed and untrusted or (2) well-formed, reasonable and trusted.
\begin{lemma}[FTLR for components]
  \label{lem:ftlr-comps}
  If $\comp$ is a well-formed component, i.e. $\wdjud{\comp}$ and either
    $\dom(\comp.\mscode) \subseteq \ta$ and $\src{\comp}$ is a reasonable component; or
    $\dom(\comp.\mscode) \mathrel{\#} \ta$,
  then there exists a $W$ such that
  $\npair{(\src{\comp},\comp)} \in \lrcomp(W)$.
\end{lemma}

@@

\section{Discussion}
\label{sec:discussion}
% \begin{itemize}
% \item explain how fully abstract overlay semantics could form one pass of a verified secure compiler.
% \item Sharing stack references accross component boundaries is supported
% \item Other notions of well-bracketedness (specifically one would be to allow different stacks)
% \end{itemize}
\subsection{Full Abstraction}
% - Full abstraction proofs difficult
Our formulation of WBCF and LSE using a fully abstract overlay semantics has an important advantage with respect to others.
Imagine that you are implementing a fully abstract compiler for a high-level language, i.e.\ a secure compiler that enforces high-level abstractions when interacting with untrusted target-language components.
Such a compiler would need to perform many things and enforce other high-level properties than just WBCF and LSE.

@@

%\subsection{Sharing the stack}
% ?
\subsection{Practical Applicability}
We believe there are good arguments for practical applicability of \stktokens{}.
The strong security guarantees are proven in a way that is reusable as part of a bigger proof of compiler security.
Its costs are
\begin{itemize}
\item a constant and limited amount of checks on every boundary crossing.
\item possibly a small memory overhead because stack frames must be of non-zero length
\end{itemize}
The main caveat is that we rely on the assumption that capability machines like CHERI can be extended with linear capabilities in an efficient way.
